---
title: "Neuroimaging in R"
author: "John Muschelli"
date: "February 19, 2015"
output:
  ioslides_presentation:
    fig_width: 4
    fig_height: 4
---

```{r knit-setup, echo=FALSE, results='hide', eval=TRUE}
rm(list=ls())
library(knitr)
setwd("~/Dropbox/Presentations/Neuroimaging_in_R/")
options(fsl.path = "/usr/local/fsl/")
opts_chunk$set(cache = TRUE, comment = "")
hook1 <- function(x){ gsub("```\n*```r*\n*", "", x) }
hook2 <- function(x){ gsub("```\n+```\n", "", x) }
knit_hooks$set(document = hook2)
```

## Introduction: Packages

- oro.nifti: read/write data, the nifti object
- fslr: process data (need FSL for most of the fun)
- ANTsR: process data (need cmake)
- extrantsr: makes ANTsR easier (need ANTsR)

## Install these packages

```{r, eval=FALSE}
install.packages("oro.nifti")
install.packages("devtools")
devtools::install_github("stnava/ANTsR") # takes long
devtools::install_github("muschellij2/extrantsr")
devtools::install_github("muschellij2/fslr")
```


## Files

Multi-modal dataset from SuBLIME (thanks Elizabeth), raw and unprocessed (other than NIfTI conversion).  

- 4 modalities: T1, T2, PD, FLAIR
- 2 time points Baseline and Followup

```{r, echo=FALSE}
files = c(base_t1="01-Baseline_T1.nii.gz", 
          base_t2="01-Baseline_T2.nii.gz",
          base_pd="01-Baseline_PD.nii.gz", 
          base_flair="01-Baseline_FLAIR.nii.gz", 
          f_t1= "01-Followup_T1.nii.gz", 
          f_t2= "01-Followup_T2.nii.gz", 
          f_pd = "01-Followup_PD.nii.gz", 
          f_flair="01-Followup_FLAIR.nii.gz")
```

```{r}
files
```

## Read in the Files!

Let's read in the baseline T1.  The `readNIfTI` function is the workhorse:
```{r read_t1, message=FALSE}
library(oro.nifti)
base_t1 = readNIfTI(files["base_t1"])
```


```{r read_t2_err, error=TRUE}
base_t2 = readNIfTI(files["base_t2"])
```

```{r read_t2}
base_t2 = readNIfTI(files["base_t2"], reorient = FALSE)
```

## An aside on `reorient` in `readNIfTI`

Arguments for `readNIfTI`:
```{r}
args(readNIfTI)
```
The default is `reorient = TRUE` but this fails **a lot**, especially when the transformation matrix is not diagonal.

##  {.flexbox .vcenter}
<div style='font-size: 35pt;'>
Use ``reorient=FALSE``
</div>

## Quick Plots

```{r ortho2, message=FALSE}
orthographic(base_t1)
```

## Quick Plots

```{r ortho, message=FALSE}
library(fslr)
fslr::ortho2(base_t1, mfrow= c(1, 3), add.orient = FALSE)
```

## Creating new nifti objects

''One image'' operations return `nifti` objects:
```{r , message=FALSE}
arr = base_t1 >= 40
class(arr)
```

## Creating new nifti objects: Copy and Replace

''Two image'' operations return `array`.
Copy `nifti` object, replace data:
```{r bad_cal_img, message=FALSE}
arr = base_t1 >= 40 & base_t1 < 60
class(arr)
mask = base_t1
mask@.Data = arr
class(mask)
```


## Creating new nifti objects: Copy and Replace

```{r bad_cal_img_plot, message=FALSE}
ortho2(mask)
```

## Creating new nifti objects: Copy and Replace

Must reset the `cal_min` and `cal_max` values, use the `fslr` function `cal_img`:

```{r cal_img, message=FALSE}
mask = cal_img(mask)
ortho2(mask)
```

## Use `cal_img` with `writeNIfTI`

`cal_img` is also useful for writing `nifti` objects.

If the range of the data isn't `c(cal_min, cal_max)`, then the `writeNIfTI` errors:
```{r, error = TRUE}
mask@.Data[5] = 3
writeNIfTI(mask, filename = tempfile())
```


## Creating new nifti objects: `niftiarr`

Use `niftiarr` to create new `nifti` objects from old one:
```{r niftiarr, message=FALSE}
mask = niftiarr(base_t1, base_t1 >= 40 & base_t1 < 60)
class(mask)
```

## Creating new nifti objects: `niftiarr`

```{r niftiarr2, message=FALSE}
ortho2(mask)
```

## Skull Stripping

Using FSL BET for the baseline T1 image:

```{r bet}
library(fslr)
ss_t1 = fslbet(base_t1, retimg = TRUE, opts = "-f 0.1 -v", 
               betcmd = "bet2", outfile = "SS_Image")
fslbin(file = "SS_Image", outfile = "Brain_Mask")
```

- `fslbet` - the command
- `retimg` - Return an image
- `opts` - options of bash `bet` command
- `betcmd` - can use `bet2` or `bet` (wrapper for `bet2`)

## Overlaying

`orthographic` allows for overlays, but must set voxels to `NA` we don't wish to see (*NOT ZERO*):
```{r overlay, message=FALSE}
library(scales)
mask = cal_img(ss_t1 > 0)
mask[mask == 0] = NA
ortho2(base_t1, y=mask, col.y=alpha("red", 0.5))
```

## Bias Field Correction

`bias_correct` from `extrantsr` package:
```{r}
library(ANTsR)
library(extrantsr)
n3_t1 = bias_correct(file = base_t1, correction = "N3", retimg=TRUE)
n4_t1 = bias_correct(file = base_t1, correction = "N4", retimg=TRUE)
```


## Registration

`extrantsr::ants_regwrite` will take in a file and a template, registre the file to the template, and can perform the reverse operation on another image.

```{r, eval=FALSE}
reg_base_t2 = ants_regwrite(filename = files["base_t2"],  
                            template.file = files['base_t1'],
                            retimg= TRUE,
                            typeofTransform = "Rigid", 
                            remove.warp = TRUE)
```


## Preprocess MRI: within a visit

`extrantsr::preprocess_mri_within` will do N3/N4 Correction, skull strip (or mask), and register to the first scan.

```{r, eval=FALSE}
proc_images = preprocess_mri_within(
  files = files[c("base_t1", "base_t2", "base_pd", "base_flair")],
    retimg = TRUE, maskfile = "Brain_Mask.nii.gz")
```

## Preprocess MRI: across visits

`preprocess_mri_across` combines `preprocess_mri_within` and `ants_regwrite`
```{r, eval=FALSE}
outfiles = gsub("[.]nii", '_process.nii', files)
preprocess_mri_across(
  baseline_files = files[c("base_t1", "base_t2", "base_pd", "base_flair")],
    followup_files = files[c("f_t1", "f_t2", "f_pd", "f_flair")],
    baseline_outfiles = outfiles[c("base_t1", "base_t2", "base_pd", "base_flair")],
    followup_outfiles = outfiles[c("f_t1", "f_t2", "f_pd", "f_flair")],
  maskfile = "Brain_Mask.nii.gz")
```


## Other useful functions from fslr

- `window_img` - set min and max values of image (good for plotting)
- `drop_img_dim` - drop dimension if is dimension is 1
- `mask_img` - mask image with an array/nifti object
- `robust_window` - quantile the image then use `window_img` (take out large values)
- `fslhd` and `fslval` can get information from an image without having to read it in
- `checkimg` - checks if an image is character, writes `nifti` to tempfile, returns character
- `check_nifti` - checks if an image is `nifti`, reads filename, returns `nifti`

## Overview

1. `oro.nifti` - reading/writing nifti objects 
2. `ANTsR`
    + bias field corrrection
    + registration: linear, non-linear (symmetric)
3. `fslr`
    + can perform image operations from FSL
    + has helper functions for `nifti` objects
4. extrantsr - helper functions for ANTsR
    + wrapper functions for preprocessing

## Bonus: Cutting Image function

```{r cut_img}
cut_img = function(img, breaks, cols, include.lowest = TRUE, plot.image=TRUE, ...){
  cuts = cut(x=c(img), breaks=breaks, include.lowest = include.lowest)
  cut_img = niftiarr(img, as.numeric(cuts))
  ybreaks= sort(unique(as.numeric(cuts)))
  ybreaks = c(0, ybreaks[!is.na(ybreaks)])
  lab = levels(cuts)
  if (plot.image){
    ortho2(img, cut_img, ybreaks = ybreaks, col.y = cols, 
         legend = lab, 
         leg.col = cols,
         addlegend = TRUE, leg.cex = 1, leg.y= 50,
         ... )
  }
  return(list(cut_img = cut_img, ybreaks= ybreaks, legend = lab, cols= cols))
}
```

## Bonus: Cutting Image function

Creating the quantiles of the skull-stripped image:
```{r cut_img_make}
img = robust_window(ss_t1)
q = quantile(c(img[img > 0]), prob = seq(0, 1, by=0.25))
breaks = unique(q)
cols = alpha(c("blue", "yellow", "orange", "red"), 0.5)
```

## Bonus: Cutting Image function
```{r cut_img_plot, dependson="cut_img"}
cimg = cut_img(ss_t1, breaks= breaks, cols = cols)
```

## Dropping out top quantile:

```{r cut_img_plot2, dependson="cut_img"}
cimg2 = cimg$cut_img
cimg2[ cimg2 == 4] = NA
cimg2 = cal_img(cimg2)
ortho2(img, cimg2, ybreaks = cimg$ybreaks, col.y = cimg$cols, 
       legend = cimg$legend, 
       leg.col = cimg$cols)
```

## Dropping out lower quantile:

```{r cut_img_plot3, dependson="cut_img"}
cimg2 = cimg$cut_img
cimg2[ cimg2 == 1] = NA
cimg2 = cal_img(cimg2)
ortho2(img, cimg2, ybreaks = cimg$ybreaks, col.y = cimg$cols, 
       legend = cimg$legend, 
       leg.col = cimg$cols)
```
